---
title: "Analysis of Factors Affecting Sleep Quality"
subtitle: "A non-parametric statistical approach"
author: "Gonçalo Araújo"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
  word_document:
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
mainfont: "Times New Roman"
fontsize: 11pt
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include= FALSE}
knitr::opts_chunk$set(echo = TRUE, width = 70)
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(knitr.include.path = FALSE)
```

\newpage

# [**Introduction**]{.underline} {#introduction}

Sleep quality is a fundamental determinant of overall health and
well-being. It is influenced by a complex interplay of psychological,
physiological, and lifestyle factors. Among these, stress levels, the
presence of sleep disorders, and engagement in physical activity are
often highlighted as critical contributors. Stress is widely recognized
for its potential to disrupt sleep patterns, leading to difficulties in
falling asleep, maintaining sleep, or achieving restorative rest.
Similarly, sleep disorders, such as insomnia or sleep apnea, directly
impair sleep quality, posing risks to both physical and mental health.
On the other hand, regular physical activity is generally associated
with better sleep outcomes, promoting deeper and more sustained sleep
cycles. Additionally, emerging research indicates potential
gender-specific variations in sleep quality, suggesting that males and
females may experience different sleep outcomes under similar
conditions. Understanding these gender-based differences could have
important implications for personalized sleep interventions and
treatments.

## Non-parametric tests

Given the critical role of sleep in health maintenance and disease
prevention, this study aims to examine the relative contributions of
stress levels, sleep disorders, physical activity, and gender on sleep
quality through non-parametric statistical analysis. This approach
allows for a robust examination of these relationships without assuming
normal distribution of the data, thereby providing more reliable
insights into these complex interactions. By computing tests including
the Mann-Whitney U test, Kruskal-Wallis test, and Spearman correlation
test, this analysis aims to infer whether these variables have
significant effects on sleep quality and to what extent they contribute
to variations in sleep outcomes.

### [Spearman Correlation Test]{.underline}

The Spearman correlation test measures the strength and direction of a
monotonic relationship between two variables. Unlike the Pearson
correlation, Spearman’s does not assume linearity or normality. Instead,
it evaluates the rank-order relationship between variables, making it
suitable for data that do not meet parametric assumptions.

The calculation involves ranking the data for both variables and
computing a correlation coefficient, based on the ranked values. This
coefficient, $r_s$ , ranges from -1 to 1, where values closer to -1 or 1
indicate a stronger negative or positive association, respectively, and
a value near 0 implies no correlation. It is calculated using the
formula:

$$
r_s = 1 - \frac{6 \sum d_i^2}{n(n^2 - 1)}
$$

Where:\
- $d_i$: The difference between the ranks of $X$ and $Y$ for each
observation ($d_i = R(x_i) - R(y_i)$).\
- $n$: The number of paired observations.

### [Kruskal-Wallis Test]{.underline}

The Kruskal-Wallis test is a rank-based non-parametric method used to
compare medians across three or more independent groups. It serves as an
alternative to the parametric one-way ANOVA when the assumptions of
normality and homoscedasticity (equal variances) are violated.

The test assesses whether the samples originate from populations with
the same median. It ranks all observations across the groups, then
computes a test statistic based on the sum of ranks within each group.
If the calculated statistic is large, it suggests significant
differences among groups. In this situations, a post hoc analysis (such
as Dunn's test) is required to determine specific group differences.

The test statistic, $H$, is calculated as:

$$
H = \frac{12}{n(n + 1)} \sum_{i=1}^{k} \frac{R_i^2}{n_i} - 3(n + 1)
$$

Where: - $n$: Total number of observations. - $k$: Number of groups. -
$R_i$: Sum of ranks for group $i$. - $n_i$: Number of observations in
group $i$.

Under the null hypothesis ($H_0$), $H$ approximately follows a
chi-squared ($\chi^2$) distribution with $k - 1$ degrees of freedom for
large sample sizes.

### [Mann-Whitney U test]{.underline}

The Mann-Whitney U test is a non-parametric alternative to the
independent samples $t$-test, designed to compare the medians of two
independent groups. It evaluates whether the distributions of two groups
differ, often used when the assumptions of normality or equal variance
are not met.

The test ranks all observations, regardless of group, and calculates the
sum of ranks for each group. The test statistic, $U$, assesses whether
one group tends to have higher or lower ranks than the other.

The test statistic $U$ is calculated using the formula:

$$
U = n_1 n_2 + \frac{n_1 (n_1 + 1)}{2} - R_1
$$

Where: - $n_1, n_2$: Sizes of the two groups. - $R_1$: Sum of ranks in
group 1.

And for group 2:

$$
U' = n_1 n_2 - U
$$

The smaller of $U$ and $U'$ is used as the test statistic.

# [**Exploratory data analysis**]{.underline}

The dataset used was the "Sleep Health and Lifestyle Dataset," available
on Kaggle. This dataset contains 13 columns covering a broad range of
variables related to sleep and daily habits. Out of the 13 available
variables, only six were selected for this analysis:

-   **Quality of Sleep:** The variable of interest, representing a
    subjective rating of sleep quality on a scale from 1 to 10.

-   **Sleep Duration:** A continuous variable indicating the number of
    hours a person sleeps per day.

-   **Physical Activity Level:** A continuous variable measuring the
    number of minutes a person engages in physical activity daily.

-   **Gender:** A binary variable indicating the gender of the
    individual (Male/Female).

-   **Stress Level:** A categorical variable with three levels:
    "Insomnia," "Sleep Apnea," and "None," representing the presence or
    absence of a sleep disorder in the individual.

```{r include=FALSE}

library(readxl)

Sleep_health_and_lifestyle_dataset <- read_excel("C:/Users/gonca/Downloads/Sleep_health_and_lifestyle_dataset.xlsx")

sleep_dataset <- Sleep_health_and_lifestyle_dataset
columns_to_remove <- c("Person ID","Age", "BMI Category", "Blood Pressure", "Heart Rate", "Daily Steps")

sleep_dataset <- sleep_dataset[, !(names(sleep_dataset) %in% columns_to_remove)]

```

## Quality of sleep vs continuous variables

```{r echo=FALSE, fig.width=8, fig.height=3, fig.align='center'}

library(ggplot2)
library(grid)
library(gridExtra)

biv_plot1 <- ggplot(sleep_dataset, aes(x = `Stress Level`, y = `Quality of Sleep`)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Quality of sleep vs Stress level",
       x = "Stress level",
       y = "Quality of sleep") +
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))

biv_plot2 <- ggplot(sleep_dataset, aes(x = `Physical Activity Level`, y = `Quality of Sleep`)) +
  geom_point(color = "purple", alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Quality of sleep vs Physical activity level",
       x = "Physical activity level",
       y = "Quality of sleep") +
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))

grid.arrange(biv_plot1,biv_plot2, ncol=2)
```

In the graph on the left, it is observable a clear negative correlation:
as stress levels increase, sleep quality tends to decrease . On the
other hand, in the graph on the right there's a slight positive
correlation between physical activity and sleep quality, though the
relationship appears weak.

## Quality of sleep vs categorical variables

```{r echo=FALSE, fig.width=8, fig.height=3, fig.align='center'}

boxplot_disorder <- ggplot(sleep_dataset, aes(x = `Sleep Disorder`, y = `Quality of Sleep`, fill = `Sleep Disorder`)) +
  geom_boxplot(show.legend = FALSE, 
               outlier.shape = NA,     
               notch = FALSE,          
               color = "black",        
               size = 0.5,             
               fatten = 2) +           
  labs(
    title = "Quality of Sleep by Sleep Disorder",
    x = "Sleep Disorder",
    y = "Quality of Sleep"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )

boxplot_gender <- ggplot(sleep_dataset, aes(x = `Gender`, y = `Quality of Sleep`, fill = `Gender`)) +
  geom_boxplot() + 
  labs(
    title = "Quality of Sleep by Gender",
    x = "Gender",
    y = "Quality of Sleep",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "none"
  )

grid.arrange(boxplot_disorder,boxplot_gender, ncol=2)
```

In the first plot, it is shown that people with no sleep disorders
generally show better sleep quality than those with insomnia, while
sleep apnea patients show the widest range of sleep quality scores.
While, in the second it is observable that females tend to report
slightly higher median sleep quality than males.

# [**Statistical analysis**]{.underline}

## Quality of sleep vs sleep duration, physical activity, stress level

```{r echo=FALSE, fig.width=8, fig.height=3, fig.align='center'}

sleep_dataset$`Quality of Sleep` <- as.numeric(sleep_dataset$`Quality of Sleep`)
correlation_sleep_duration <- cor.test(
  sleep_dataset$`Quality of Sleep`, 
  as.numeric(sleep_dataset$`Sleep Duration`), 
  method = "spearman"
)
correlation_physical_activity <- cor.test(
  sleep_dataset$`Quality of Sleep`, 
  as.numeric(sleep_dataset$`Physical Activity Level`), 
  method = "spearman"
)
correlation_stress_level <- cor.test(
  sleep_dataset$`Quality of Sleep`, 
  as.numeric(sleep_dataset$`Stress Level`), 
  method = "spearman"
)

correlation_results <- data.frame(
  Variable = c("Sleep Duration", "Physical Activity Level", "Stress Level"),
  Correlation = c(
    correlation_sleep_duration$estimate,
    correlation_physical_activity$estimate,
    correlation_stress_level$estimate
  )
)

heatmap_plot <- ggplot(correlation_results, aes(x = Variable, y = "Quality of Sleep", fill = Correlation)) +
  geom_tile(color = "white", width = 0.6, height = 0.6) +  
  geom_text(aes(label = round(Correlation, 2)), size = 3, color = "black") +  
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limit = c(-1, 1),
    name = "Spearman\nCorrelation"
  ) +
  theme_minimal() +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Correlation with Quality of Sleep",
    x = NULL
  ) +
  coord_fixed()

print(heatmap_plot)
```

The **Spearman rank correlation test** was conducted to assess the
relationship between quality of sleep and several variables, including
sleep duration, physical activity level, and stress levels.

-   **Quality of Sleep vs. Sleep Duration:** A strong positive
    correlation was found between quality of sleep and sleep duration
    (rho = 0.89, *p* \< 2.2e-16). This suggests that individuals who
    reported higher quality of sleep tend to also have longer sleep
    durations, with the correlation being highly significant.

-   **Quality of Sleep vs. Physical Activity Level:** A weak positive
    correlation was observed between quality of sleep and physical
    activity level (rho = 0.18, *p* = 0.0005249). While the correlation
    is positive, it is relatively low, indicating that physical
    activity may have a small positive effect on sleep quality.

-   **Quality of Sleep vs. Stress Level:** A very strong negative
    correlation was found between quality of sleep and stress level (rho
    = -0.91, *p* \< 2.2e-16). This result suggests that higher stress
    levels are strongly associated with poorer sleep quality, supporting
    the idea that stress negatively impacts sleep.

Overall, these findings highlight significant relationships between
quality of sleep and both sleep duration and stress levels, while
physical activity level showed a weaker, yet still significant, light
positive association.

## Comparison of quality of sleep across sleep disorders

The **Kruskal-Wallis test** was performed to assess whether the quality
of sleep differs across groups of individuals with different sleep
disorders. The analysis revealed a significant result (Kruskal-Wallis
test = 49.179, df = 2, *p*-value = 2.094e-11), indicating that there are
differences in the quality of sleep between the groups (Insomnia, Sleep
Apnea, and None). Following the Kruskal-Wallis test, post-hoc
comparisons using Dunn's test were conducted to identify which specific
groups differ from one another. The results of the Dunn's test showed
the following significant pairwise differences:

**Insomnia vs. None:** The difference between the groups was highly
significant (Z = -6.9977, *p*-value = 1.30e-12).

**Insomnia vs. Sleep Apnea:** The difference between these groups was
also highly significant (Z = -3.9057, *p*-value = 4.70e-5).

**None vs. Sleep Apnea:** A significant difference was found between
these two groups (Z = 2.2727, *p*-value = 0.0115).

The *p*-values were adjusted for multiple comparisons using the
Bonferroni correction. After adjustment, the *p*-values remained
significant for Insomnia vs. None and Insomnia vs. Sleep Apnea, and
marginally significant for None vs. Sleep Apnea (adjusted p-value =
0.0345).

The results suggest that individuals with Insomnia report significantly
worse sleep quality compared to those with "None" and those with "Sleep
Apnea". Overall, these findings provide important insights into the
relationship between different sleep disorders and their impact on sleep
quality.

## Comparison of sleep quality by gender

A **Mann-Whitney U test** was performed to compare the quality of sleep
between males and females. The results indicated a significant
difference between the two groups (W = 23369, *p*-value = 6.048e-09).
This suggests that the distribution of quality of sleep differs
significantly between genders.

The median quality of sleep for females was 8, while for males it was 7,
indicating that females tend to report slightly better sleep quality on
average than males. To further understand the magnitude of this
difference, an effect size (r) for the Mann-Whitney test was calculated.
The effect size (which quantifies the strength of the difference between
the two groups) was found to be 0.3007, indicating an average effect.

In summary, these results provide evidence that females report better
quality of sleep compared to males, with a medium effect size indicating
the practical significance of this difference.

# [**Conclusions**]{.underline}

This study explored the impact of several factors on quality of sleep.
Key findings of this work include:

-   **Sleep Duration:** Longer sleep is closely linked to better sleep
    quality.

-   **Physical Activity:** While physical activity is positively
    associated with sleep quality, its effect is relatively modest
    compared to other factors.

-   **Stress Levels:** Higher stress levels have a significantly
    negative impact on sleep quality.

-   **Sleep Disorders:** Individuals with Insomnia reported the poorest
    sleep quality, followed by those with Sleep Apnea. People without
    sleep disorders generally reported the best sleep quality,
    indicating that addressing sleep disorders is essential for
    improving sleep health.

-   **Gender:** Females generally reported better sleep quality than
    males, suggesting that gender may influence sleep patterns and
    perceptions of sleep quality.

Overall, sleep duration, stress management, and addressing sleep
disorders are the most significant factors influencing sleep quality.
Gender also plays a role, with females typically reporting better sleep
quality than males.

# [References]{.underline}

1.  Alwhaibi, M., & Al Aloola, N. A. (2023). Associations between
    Stress, Anxiety, Depression and Sleep Quality among Healthcare
    Students. *Journal of clinical medicine*, *12*(13), 4340.
    https://doi.org/10.3390/jcm12134340

2.  Alnawwar, M. A., Alraddadi, M. I., Algethmi, R. A., Salem, G. A.,
    Salem, M. A., & Alharbi, A. A. (2023). The Effect of Physical
    Activity on Sleep Quality and Sleep Disorder: A Systematic
    Review. *Cureus*, *15*(8), e43595.
    https://doi.org/10.7759/cureus.43595

3.  Kohanmoo, A., Kazemi, A., Zare, M. *et al.* Gender-specific link
    between sleep quality and body composition components: a
    cross-sectional study on the elderly. *Sci Rep* **14**, 8113 (2024).
    https://doi.org/10.1038/s41598-024-58801-5

4.  Riffenburgh, R. H. (2006). Chapter summaries. In R. H. Riffenburgh
    (Ed.), *Statistics in medicine* (2nd ed., pp. 533–580). Academic
    Press. https://doi.org/10.1016/B978-012088770-5/50067-8

5.  Dinno, A. (2024). *dunn.test: Dunn's Test of Multiple Comparisons
    Using Rank Sums* (Version 1.3.6) [R package]. Comprehensive R
    Archive Network (CRAN).\
    https://doi.org/10.32614/CRAN.package.dunn.test

6.  Zar, J. H. (2010). *Biostatistical analysis* (5th ed.). Pearson
    Education.

7.  Conover, W. J. (1999). *Practical nonparametric statistics* (3rd
    ed.). Wiley. https://doi.org/10.1002/9781118032915

8.  Hollander, M., Wolfe, D. A., & Chicken, E. (2014). *Nonparametric
    statistical methods* (3rd ed.). Wiley.
    https://doi.org/10.1002/9781119196036
